Communicative pressures at the semantics-pragmatics interface:\\ Learning biases may prevent the lexicalization of pragmatic inferences

### 2016-09-26 paper/paper.tex
new labels and style for paper plots + Added more to the analysis section

\subsection{Mutation \& learning}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type to adopt from the resulting posterior distribution.

@@

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation. If not stated otherwise, the results for a given parameter setting were obtained from $1000$ independent runs. Each run consisted of $20$ generations. This corresponds to a developmental plateau after which no noteworthy change was registered. The model's parameters are summarized in Table \ref{tab:summary}.

@@

\hl{MENTION SAMPLING FROM PRODUCTION}

According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. However, it is instructive to first inspect the effect of these pressures in isolation. We will focus three pragmatic types.\footnote{Pragmatic reasoning allows hearers and speakers to refine their (possibly erroneous) interpretations, rendering it advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds. Users of $L_4$, a type that is functionally advantageous but biased against. And users of $L_5$, combining the virtues of the latter two.

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{./only-R-or-M}
\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (left; $\alpha = 1$) and only mutation (right; $\lambda = 20, k = 5$).}
\label{fig:either-R-or-M}
\end{figure}

\paragraph{Learnability only.} To get an impression of the effect of iterated learning without a pressure towards expressivity, we first consider (relatively) deterministic parental data production ($\lambda = 20, \alpha = 1$). In this way small data sequences suffice for learners to differentiate types that produce strongly diverging expressions. In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior ($l = 1$). This is shown in the right-hand side of Figure \ref{fig:either-R-or-M}, which directly reflects the prior distribution for each value of $c$.

Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some further-reaching issues in our setup. First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations. For instance, while types using $L_4$ can come to populate a large proportion of the final population, a very low degree of rationality needs to be assumed.

\hl{How about IL with higher L? Take a look at the prior please!}

\begin{itemize}
  \item Discuss RMD;
  \item Discuss prior and posterior;
\end{itemize}

\begin{figure}
\centering
\includegraphics[scale=.5]{../presentations/03cost-with-l1}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.5]{../presentations/01heatmap}
\end{figure}

\paragraph{Results.} We report the mean results obtained from $1000$ independent simulations for each set of parameters. Figure \ref{fig:c-85} illustrates the development of the target types over $20$ generations with $c = 0.85$, i.e., a strong bias against semantic complexity. Figure \ref{fig:cost} shows the effect of the learning bias after $20$ generations for $c \in [0,1]$. More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}.

\begin{table}
\centering 
\begin{tabular}{l c c c c c c c c c c c c c}
~ & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
c  & 0 & .01 & .05 & .1 & .8 & empty & 0 & .01 & .05 & .1 & .8\\ \hline \hline
lit. $L_1$& ~ & ~ & ~ & ~  & ~& ~ & ~& ~ & ~ & ~ & ~\\ 
lit. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
lit. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
lit. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
lit. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\
lit. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~\\ \hline
prg. $L_1$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
prg. $L_2$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
prg. $L_3$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\ 
prg. $L_4$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
prg. $L_5$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~ \\
prg. $L_6$ & ~ & ~ & ~ & ~ & ~& ~ & ~& ~ & ~ & ~ & ~

\end{tabular}
\caption{Mean proportion of types in $1000$ populations after $20$ generations across different bias values $c$ ($\alpha = 1, \lambda = 50, k = 20$).}

\label{tab:numeric-results}
\end{table}

In sum, the results show that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$. This effect increases with the bias' strength provided $L_5$ users are pragmatic. That is, such a learning bias may prevent the lexicalization of pragmatic inferences and explain the prevalence of $L_5$-like semantics. However, note that literal $L_5$ is underrepresented across all values of $c$. This highlights that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, in itself it does not enable types that fail to convey an upper-bound to establish themselves in the population. This suggests pragmatic reasoning to be paramount to the selection of $L_5$-like semantics. That is, neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.

>>>>

\subsection{Mutation \& learning}\label{sec:mutation}
Iterated learning is a process in which the behavior of one individual serves as learning input for another, who's behavior subsequently serves as input for a new learner, and so on. For linguistic purposes this process can be thought of as chains of parents and children, where the parent produces linguistic data from which the child infers a language. The latter, now a parent, goes on to produce linguistic data for a new generation of na\"ive learners. Following \citet{griffiths+kalish:2007} we model learning as a process of Bayesian inference in which learners combine the likelihood of a type producing the learning data with prior inductive biases. They then select a type to adopt from the resulting posterior distribution.

@@

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting the population's first generation. If not stated otherwise, the results for a given parameter setting were obtained from $1000$ independent runs. Each run consisted of $20$ generations. This corresponds to a developmental plateau after which no noteworthy change was registered. As specified in \ref{sec:mutation}, the learning transition matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, matrices with $k > 5$ were approximated by using Monte Carlo with $10$ sequences sampled from each type's production probabilities and a type's children being exposed only to this subset. The model's parameters are summarized in Table \ref{tab:summary}.

@@

According to our hypothesis, functional pressure on successful communication combined with learning pressures in the form of a bias against upper-bounds may lead to the selection of $L_5$-like semantics. However, it is instructive to first inspect the effect of these pressures in isolation. For this purpose, we focus on three pragmatic types.\footnote{Pragmatic reasoning allows language users to refine their (possibly erroneous) choices. Therefore, it is advantageous even for those types that codify an upper-bound lexically.} Users of $L_3$, representing a type that is lacking in expressivity but is a prior preferred for its lack of upper-bounds. Users of $L_4$, a type that is functionally advantageous but biased against. And users of $L_5$, combining the virtues of the latter two.

@@

\begin{figure}
\centering
\includegraphics[scale=.5]{./only-R-or-M}
\caption{Mean proportions of target types after $20$ generations in $1000$ populations with only replication (A; $\alpha = 1$) and only mutation (B; $\alpha =1, \lambda = 30, k = 5$).}
\label{fig:either-R-or-M}
\end{figure}

\paragraph{Learnability only.} To get an impression of the effect of iterated learning without a pressure towards expressivity, we first consider (relatively) deterministic parental data production ($\lambda = 20, \alpha = 1$). In this way small data sequences suffice for learners to differentiate types that produce strongly diverging expressions. In line with \citeposs{griffiths+kalish:2007} analysis, under these conditions the resulting populations converge to the learner's prior distribution when sampling from the posterior. This is shown in the right-hand side of Figure \ref{fig:either-R-or-M}, which directly reflects the prior distribution for each value of $c$.

Inspecting the effects of these dynamics separately not only gives some intuitions about the parameters' influence, but also highlights some of their broader implications. First and foremost, neither dynamic comes close to converging to a monomorphic population under most parameter configurations. For instance, types using $L_4$ can come to occupy a large proportion of the final population. However, this holds only for a restricted range of low degrees of rationality. Apart from polymorphy, both dynamics make some undesirable predictions. A pressure only towards expressivity leads to the selection of types using $L_4$ to $L_6$ and to the ejection of $L_1$ to $L_3$. However, it can not explain the regular selection of $L_5$-like semantics over either of these functionally similar alternatives. In contrast, a pressure only towards learnability has a modest but clear effect in differentiating $L_5$ from these alternatives but fails to rule out functionally suboptimal types such as tautological $L_3$. This showcases that, while the bias plays a major role for the contrast between $L_4$ and $L_5$, on its own it does not enable types that fail to convey an upper-bound to establish themselves in the population. In sum, neither dynamic on its own is a suitable candidate to provide a justification for the predicted prevalence of $L_5$-like semantics.

\paragraph{Expressivity and learnability.} Figure \ref{fig:cost} illustrates the effect of the learning bias after $20$ generations across values of $c$ with $l = 1$ (A) and $l =3$ (B). More detailed results for all types across a sample of $c$-values are presented in Table \ref{tab:numeric-results}. Overall, these results suggest that in the present setup a weak bias is sufficient to lead to a selection of $L_5$ over $L_4$. As in the simulations that only considered learnability, this effect increases with the bias' strength provided $L_5$ users are pragmatic. Importantly, the addition of a pressure towards expressivity magnifies this effect by dampening the proliferation of functionally suboptimal types advantaged by the learning bias. As stressed above, this suggests that neither the learning bias nor functional pressure alone but their combination lead to the systematic lack of semantic upper-bounds in scalar expressions.

\begin{figure}
\centering
\includegraphics[scale=.5]{./fig2-rmd}
\label{fig:cost}
\caption{Mean proportions of target types after $20$ generations in $1000$ populations across bias values $c \in [0,1]$ with $l =1$ in A and $l = 3$ in B ($\alpha =1, \lambda = 20, k = 5$).}
\end{figure}

\begin{table}
\centering 
\begin{tabular}{l | c c c c c| c c c c c c c c}
\multicolumn{1}{c}{~} & \multicolumn{5}{c}{$l = 1$} & ~ & \multicolumn{5}{c}{$l = 10$}\\ \hline \hline
c  & 0 & .01 & .05 & .1 & .8 & ~ & 0 & .01 & .05 & .1 & .8\\ \hline \hline
lit. $L_1$& .03 & .03 & .03 & .03  & .04 & ~ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ 
lit. $L_2$ & .03 & .03 & .03 & .03 & .01 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
lit. $L_3$ & .03 & .03 & .03 & .03 & .04 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
lit. $L_4$ & .07 & .07 & .07 & .07 & .06 & ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
lit. $L_5$ & .04 & .04 & .05 & .05 & .06& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\
lit. $L_6$ & .04 & .04 & .04 & .04 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$\\ \hline
prg. $L_1$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
prg. $L_2$ & .03 & .03 & .03 & .03 & .01& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\
prg. $L_3$ & .03 & .03 & .03 & .03 & .04& ~ & $\epsilon$& $\epsilon$ & $\epsilon$ & $\epsilon$ & $\epsilon$ \\ 
prg. $L_4$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15 \\
prg. $L_5$ & .22 & .22 & .22 & .23 & .3& ~ & .33& .33 & .35 & .37 & .7 \\
prg. $L_6$ & .22 & .22 & .22 & .22 & .18& ~ & .33& .33 & .32 & .31 & .15   
\end{tabular}
\caption{Mean proportions of types in $1000$ populations after $20$ generations across bias values $c \in [0,1]$ with $l =1$ and $l = 3$ ($\alpha = 1, \lambda = 30, k = 5$), $\epsilon < 0.005$}
\label{tab:numeric-results}
\end{table}

This data suggests that the proportion of scalar implicature users that is predicted primarily hinges on three aspects of the model. First, the degree to which linguistic behavior is deterministic, controlled by $\lambda$ and $\alpha$, plays a role both for expressivity as well as for producing data that allows learners to discriminate this type from others. Second, the learning bias $c$, leads learners to discriminate and prefer $L_5$ over $L_4$ and $L_5$. Lastly, the posterior parameter $l$ magnifies the effects of the learning bias in tandem with replication. This interaction is shown in \ref{fig:prior-posterior}. In the present setup posterior sampling can lead to the incumbency of pragmatic $L_5$, but not even a strong favorable learning bias manages to completely drive out competing types (cf. \ref{fig:cost}.A). However, as posterior maximization increases, the range of bias values within which this type takes over the population increases drastically.

\hl{More  discussion about parameter interaction}

\begin{figure}
\centering
\includegraphics[scale=.5]{../presentations/01heatmap}
\caption{Mean proportion of pragmatic $L_5$ in $1000$ populations after $20$ generations ($\alpha = 1, \lambda = 30, k = 5$)}
\label{fig:prior-posterior}
\end{figure}
